Undecidability and the structure of the Turing degrees

### 2018-08-14 paper.tex
jump operator definition

% -------Packages---------
\usepackage{microtype}[final]
\usepackage[style=numeric]{biblatex}
\usepackage{hyperref}
\addbibresource{/home/philip/documents/text/references/bibliography.bib}
\usepackage{amssymb,amsfonts}
\usepackage[all,arc]{xy}
\usepackage{enumerate}
\usepackage{mathrsfs}
\usepackage{todonotes}
\usepackage{tikz}
\usetikzlibrary{automata,positioning}
\usepackage[nolist]{acronym}

@@

\tableofcontents
\begin{acronym}
  \acro{PDA}[PDA]{Pushdown Automaton}
  \acroplural{PDA}[PDAs]{Pushdown Automata}
  \acro{DPDA}[DPDA]{Deterministic Pushdown Automaton}
  \acroplural{DPDA}[DPDAs]{Deterministic Pushdown Automata}
  \acro{TM}[TM]{Turing Machine}
  \acro{FSA}[FSA]{Finite State Automaton}
  \acroplural{FSA}[FSAs]{Finite State Automata}
  \acro{CFG}[CFG]{Context-Free Grammar}
  \acro{CFL}[CFL]{Context-Free Language}
\end{acronym}

\section{Introduction}
Historical overview of the field, motivation for study
\cite{ambos-spies06:_degrees_unsol}
\cite{soare16_turin_comput}
\cite{soare1999history}
\cite{lerman16:_degrees_unsol}
\subsection{Decision Problems}

@@

\begin{prob}[The Halting Problem]
  \label{prob:halting}
  Let $M$ be a Turing Machine and $i$ be an input. If $M$ is run on $i$, will it
  eventually halt?
  \begin{proof}[Proof of Undecidability]
    Suppose for the sake of contradiction that the halting problem is
    decidable. Then, there exists some machine
    $\mathcal{O}$ that can decide the halting problem. Then, we can construct a
    Turing machine $H$ that simulates $\mathcal{O}$ on its input $(M,i)$. In the
    case where $\mathcal{O}(M,i)$ accepts, $H$ enters into an infinite
    loop, whereas in the case where $\mathcal{O}(M,i)$ does not accept, $H$
    halts. Consider $H(H,\epsilon)$. If $H$ halts, then
    $\mathcal{O}(H,\epsilon)$ accepts, and so $H$ enters an infinite loop, and
    so it does not halt. But if $H$ does not halt, then
    $\mathcal{O}(H,\epsilon)$ does not accept, so $H$ halts. This is a
    contradiction, so our premise that the halting problem is decidable must be false.
  \end{proof}

  \cite{sipser13:_introd_theor_comput}
\end{prob}
\begin{thm}[Rice's Theorem] \cite{kozen99_autom}
  % Proof by reduction to the halting problem?
\end{thm}
\subsection{Incompleteness}
\begin{thm}[G\"odel's Incompleteness Theorem] 
  % Kleene 1943? Also shown in hopcroft 1979
  \cite{kleene43_recur_predic_quant}
\end{thm}
\subsection{Reducibility}
\label{subsec:reducibility}
While it is possible to prove that a problem is undecidable directly, as in
problem~\ref{prob:halting}, it is often more convenient to prove undecidability
through comparison to problems which are already known to be undecidable. This
comparison takes place through the technique of Turing reduction.
\begin{defn}
  Let $A$ and $B$ be decision problems. We say that $A$ is \emph{Turing reducible}
  to $B$ and write $A \leq_T B$ if, given a machine $D_B$ that decides $B$, it is possible to construct
  a machine $D_A$ that decides $A$.
\end{defn}
So, we have that if $A$ is reducible to $B$, then $A$ can be no harder than $B$,
because any solution to $B$ also leads to a solution of $B$. So, if we have some
problem $B$ that we would like to prove is undecidable, we can do so by showing
that some problem $A$ which is already known to be undecidable is reducible to
$B$. Since $A$ cannot be harder than $B$, it follows that $B$ must also be
undecidable. Alternatively, if we would like to show that some problem $A$ is
decidable, it is sufficient to show that it is reducible to some problem $B$
that is known to be decidable.
\cite{sipser13:_introd_theor_comput}
\cite{post44:_recur}

@@

\section{Context Free Grammars}

\begin{defn}
 A Context Free Grammar is...
\end{defn}

% \begin{defn}
%   A Pushdown Automata is...
% \end{defn}

\begin{thm}[Equivalence of CFGs and Pushdown Automata]
  \cite{hopcroft07:_introd_autom_theor_languag_comput}
\end{thm}
\begin{thm}[Closure Properties]
  \cite{sipser13:_introd_theor_comput}
  Union, Concat, Star, intersection with regular, substitution
\end{thm}
\begin{prob}[Emptiness]
  \label{prob:cfg:emptiness}
Let $G$ be a context-free grammar. Is $L(G)=\varnothing$?
\end{prob}

\begin{prob}[Finite]
Let $G$ be a context-free grammar. Is $L(G)$ finite?
\end{prob}

@@

\begin{defn}
An ambiguous grammar is...
\end{defn}

@@

\subsubsection{Problems decidable for Deterministic CFLs}
\cite{ginsburg65:_deter}
\begin{defn}
  A deterministic context-free language is a language accepted by a
  deterministic pushdown automata.
\end{defn}

\begin{thm}[Closure Properties]
  If $L$ is a deterministic context-free language, then $\overline{L}$ is a
  deterministic context-free language.
\end{thm}

\begin{prob}
  Let $L$ be a deterministic context-free language. Is $L = \Sigma^*$?
\end{prob}

@@

\begin{prob}[The Equivalence Problem for Deterministic CFGs]
  Let $L_1,L_2$ be deterministic context-free languages. Is $L_1=L_2$?

@@

\begin{defn}
  Computably Enumerable
\end{defn}

\begin{defn}
  Completeness
\end{defn}

@@

\begin{defn}
  The jump operator
\end{defn}

@@

% Myhill 56 c.e.lattice

\begin{lem}
  D does not form a lattice
  \cite{kleene54_upper_semi_lattic_degrees_recur_unsol}
\end{lem}
\begin{thm}
  D forms an upper semi-lattice
  % Kleene-Post 54 Degrees upper semi-lattice, jump an operator on degrees
  \cite{kleene54_upper_semi_lattic_degrees_recur_unsol}
\end{thm}

@@

\subsection{Turing Degrees of Problems related to CFGs}
\cite{REEDY197577}

>>>>

% -------Packages---------
\usepackage{microtype}[final]
\usepackage[style=numeric]{biblatex}
\usepackage{hyperref}
\addbibresource{/home/philip/documents/text/references/bibliography.bib}
\usepackage{amssymb,amsfonts}
\usepackage[all,arc]{xy}
\usepackage{enumerate}
\usepackage{mathrsfs}
\usepackage{todonotes}
\usepackage{tikz}
\usetikzlibrary{automata,positioning,shapes}
\usepackage[nolist]{acronym}

@@

\tableofcontents
\begin{acronym}
  \acro{PDA}[PDA]{Pushdown Automaton}
  \acroplural{PDA}[PDAs]{Pushdown Automata}
  \acro{DPDA}[DPDA]{Deterministic Pushdown Automaton}
  \acroplural{DPDA}[DPDAs]{Deterministic Pushdown Automata}
  \acro{TM}[TM]{Turing Machine}
  \acro{FSA}[FSA]{Finite State Automaton}
  \acroplural{FSA}[FSAs]{Finite State Automata}
  \acro{CFG}[CFG]{Context-Free Grammar}
  \acro{CFL}[CFL]{Context-Free Language}
  \acro{DCFL}[DCFL]{Deterministic Context-Free Language}
\end{acronym}

\section{Introduction}
Historical overview of the field, motivation for study
\cite{ambos-spies06:_degrees_unsol}
\cite{soare16_turin_comput}
\cite{soare1999history}
\cite{lerman16:_degrees_unsol}
\todo{}
\subsection{Decision Problems}

@@

\begin{prob}[The Halting Problem]
  \label{prob:halting}
  Let $M$ be a Turing Machine and $i$ be an input. If $M$ is run on $i$, will it
  eventually halt?
  \begin{proof}[Proof of Undecidability]
    Suppose for the sake of contradiction that the halting problem is
    decidable. Then, there exists some machine
    $\mathcal{O}$ that can decide the halting problem. Then, we can construct a
    Turing machine $H$ that simulates $\mathcal{O}$ on its input $(M,i)$. In the
    case where $\mathcal{O}(M,i)$ accepts, $H$ enters into an infinite
    loop, whereas in the case where $\mathcal{O}(M,i)$ does not accept, $H$
    halts. Consider $H(H,\epsilon)$. If $H$ halts, then
    $\mathcal{O}(H,\epsilon)$ accepts, and so $H$ enters an infinite loop, and
    so it does not halt. But if $H$ does not halt, then
    $\mathcal{O}(H,\epsilon)$ does not accept, so $H$ halts. This is a
    contradiction, so our premise that the halting problem is decidable must be false.
  \end{proof}
\end{prob}
% \begin{thm}[Rice's Theorem] \cite{kozen99_autom}
%   % Proof by reduction to the halting problem?
% \end{thm}
% \subsection{Incompleteness}
% \begin{thm}[G\"odel's Incompleteness Theorem] 
%   % Kleene 1943? Also shown in hopcroft 1979
%   \cite{kleene43_recur_predic_quant}
% \end{thm}
\subsection{Reducibility}
\label{subsec:reducibility}
While it is possible to prove that a problem is undecidable directly, as in
problem~\ref{prob:halting}, it is often more convenient to prove undecidability
through comparison to problems which are already known to be undecidable. This
comparison takes place through the technique of Turing reduction.
\begin{defn}
  Let $A$ and $B$ be decision problems. We say that $A$ is \emph{Turing reducible}
  to $B$ and write $A \leq_T B$ if, given a machine $D_B$ that decides $B$, it is possible to construct
  a machine $D_A$ that decides $A$.
\end{defn}
So, we have that if $A$ is reducible to $B$, then $A$ can be no harder than $B$,
because any solution to $B$ also leads to a solution of $B$. So, if we have some
problem $B$ that we would like to prove is undecidable, we can do so by showing
that some problem $A$ which is already known to be undecidable is reducible to
$B$. Since $A$ cannot be harder than $B$, it follows that $B$ must also be
undecidable. Alternatively, if we would like to show that some problem $A$ is
decidable, it is sufficient to show that it is reducible to some problem $B$
that is known to be decidable.
\cite{sipser13:_introd_theor_comput}
\cite{post44:_recur}

@@

\section{Context Free Languages}

\begin{defn}
 A \ac{CFG} is a finite set of nonterminal symbols, a finite set of terminal
 symbols, a finite number of production rules that produce a string of terminal
 and nonterminal symbols from a nonterminal symbol, and a starting symbol.
\end{defn}

\begin{defn}
  The \emph{language} of a \ac{CFG} is the set of words that can be formed by
  the successive application of production rules to the start symbol. These
  languages are called \acp{CFL}.
\end{defn}

\acp{CFG} provide another useful model of computation. However, the languages
that they are able to
express are exactly the same set of languages as \acp{PDA}~\cite{hopcroft07:_introd_autom_theor_languag_comput}. This equivalence allows
us to choose between the two models and use the one that is most useful for
proving facts about these languages.

\begin{thm}[Closure Properties of \acp{CFL}]
  \aclp{CFG} have a variety of closure properties:
  \begin{itemize}
  \item Let $L_1,L_2$ be \acp{CFL}. Then $L_1\cup L_2$ is a \ac{CFL}.
    \begin{proof}
      The languages $L_1$ and $L_2$ are \aclp{CFL}, so there exist \aclp{PDA} $P_1$ and $P_2$
      that decide them. Consider the \ac{PDA} $U$ shown in
      Figure~\ref{pda:union}, which simulates $P_1$ and $P_2$ on the input. $U$
      is a \ac{PDA} that decides $L_1\cup L_2$, so it follows that
      $L_1\cup L_2$ is a \ac{CFL}.
      \begin{figure}[h]
        \caption{$U$, a \ac{PDA} that simulates $P_1$ and $P_2$.}
        \label{pda:union}
         \begin{tikzpicture}[shorten >=1pt,node distance=3cm,on grid,auto]
  \node[state,initial] (q_0) {$q_0$};
  \node[state,diamond] (P_1) [above right=of q_0]{$P_1$};
  \node[state,diamond] (P_2) [below right=of q_0]{$P_2$};
  \draw[->]
  (q_0) edge node {$\epsilon,\epsilon\to\epsilon$} (P_1);
  \draw[->]
  (q_0)
  edge node {$\epsilon,\epsilon\to\epsilon$} (P_2);
  \end{tikzpicture}
      \end{figure}
    \end{proof}
  \item Let $L_1,L_2$ be \acp{CFL}. Then the concatenation \[
      \{w_1w_2 \mid w_1\in L_1, w_2 \in L_2\}
    \]
    is a \ac{CFL}.
    \begin{proof}
      The languages $L_1$ and $L_2$ are \aclp{CFL}, so there exist \acp{CFG}
      $G_1,G_2$ with start symbols $S_1,S_2$ that produce them. Suppose, without loss of generality, that
      $G_1,G_2$ have no nonterminal symbols in common (if they do, then we can rename
      them). Consider the grammar $S \rightarrow S_1S_2$. Then, this is a
      context free grammar, and its language is the concatenation of $L_1$ and
      $L_2$. So, the concatenation of $L_1$ and $L_2$ is a \ac{CFL}.
    \end{proof}
  \item Let $L$ be a \ac{CFL}. Then,
    \[
      L^* = \bigcup_{i \geq 0}\{w_1 \dots w_k\dots w_i \mid w_k\in L\}
    \]
    is a \ac{CFL}.
    \begin{proof}
      The language $L$ is a \ac{CFL}, so there exists a \ac{CFG} $G$ with start
      symbol $S_1$ that produces it. Consider the grammar given by $S\rightarrow
      S_1S\mid\epsilon$. This is a \ac{CFG} that produces $L^*$, so it follows
      that $L^*$ is a \ac{CFL}.
    \end{proof}
    \item Intersection with regular languages~\cite{hopcroft07:_introd_autom_theor_languag_comput}.\todo{}
  \end{itemize}
\end{thm}

\begin{prob}[Emptiness]
  \label{prob:cfg:emptiness}
  Let $G$ be a context-free grammar. Is $L(G)=\varnothing$?
  \begin{proof}[Proof of Decidability]
    \todo{}
  \end{proof}
\end{prob}

\begin{prob}[Finite]
  Let $G$ be a context-free grammar. Is $L(G)$ finite?
  \begin{proof}[Proof of Decidability]
    \todo{}
  \end{proof}
\end{prob}

@@

\begin{defn}
We say that a \acl{CFG} is \emph{ambiguous} if there is more than one way to
apply the production rules to reach an accepted word, assuming a production rule
is always applied to the leftmost nonterminal symbol. This is equivalent to
stating that there is more than one parse tree that parses an accepted word.
\end{defn}

@@

\subsubsection{Problems decidable for DCFLs}
\cite{ginsburg65:_deter}
\begin{defn}
  A \ac{DCFL} is a language accepted by a
  \acl{DPDA}.
\end{defn}

\begin{thm}[Closure Properties]
  \label{dcfl:complement}
  If $L$ is \ac{DCFL}, then $\overline{L}$ is a
  \ac{DCFL}. \todo{}
\end{thm}

\begin{prob}
  Let $L$ be a \ac{DCFL}. Is $L = \Sigma^*$?
  \begin{proof}[Proof of Decidability]
    We have that $L$ is a \ac{DCFL}, so by Theorem~\ref{dcfl:complement},
    $\overline{L}$ is also a $\ac{DCFL}$. By Problem~\ref{prob:cfg:emptiness},
    we have that whether $\overline{L}$ is empty is decidable, so it follows
    that the question of whether $L = \Sigma^*$ is also decidable.
  \end{proof}
\end{prob}

@@

\begin{prob}[The Equivalence Problem for Deterministic CFGs]
  Let $L_1,L_2$ be deterministic context-free languages. Is $L_1=L_2$?
\todo{}
\end{prob}

@@

\begin{defn}
  An oracle machine for a degree $A$, denoted $\mathcal{O}_A$, is a Turing
  machine equipped with a function that is able to decide $A$.
\end{defn}

\begin{defn}
  A problem $P$ is called \emph{computably enumerable} or \emph{semidecidable}
  for some degree $d\in\mathcal{D}$ if there exists a problem $A\in d$ such that
  it is possible to construct an oracle machine $\mathcal{O}_A$ such that
  $\mathcal{O}_A$ halts for all the inputs where $P$ is true, and does not halt
  on all the inputs for which $P$ is false. If $A=\varnothing$, and so the
  oracle machine does not have an oracle, then we simply say that $P$ is
  computably enumerable, semidecidable, or \emph{Turing recognizable}.
\end{defn}
\begin{rem}
  The Halting problem, Problem~\ref{prob:halting}, is Turing recognizable. The
  machine that recognizes the problem simply simulates the input machine.
\end{rem}

@@

\begin{defn}
  The jump of $A$, denoted $A'$, is degree of the halting problem for oracle
  machines for $A$.
\end{defn}

The jump operator has many properties that make it convenient for studying the
Turing degrees, which are listed below. Further commentary on the jump operator
and its properties is available in Robert Soare's book \emph{History of Computability}~\cite{soare16_turin_comput}.
\begin{fact}
  The jump operator has many properties that make it convenient for studying the
Turing degrees, some of which are listed below. Further commentary on the jump operator
and its properties is available in Robert Soare's book \emph{History of Computability}~\cite{soare16_turin_comput}.
  \begin{itemize}
  \item $A'$ is computably enumerable in $A$.
  \item $A' \not\leq_T A$.
  \item If $B=_TA$ then $B'=_TA'$.
  \end{itemize}
\end{fact}

@@

% Myhill 56 c.e.lattice
\begin{defn}
  A partially ordered set $S$ is said to form an \emph{upper semilattice} if, for every
 $x,y\in S$, there exists some element $z\in S$ such that $z \geq x,y$ and
 for all $w\in A$ such that $w\geq x,y$, we have that $w \geq z$. In other
 words, $x$ and $y$ have a supremum.
\end{defn}

\begin{defn}
  A partially ordered set $S$ is said to form a \emph{lattice} if all pairs
  $\{x,y\}\in S$ have both an infinum and a supremum.
\end{defn}

\begin{thm}
  $\mathcal{D}$ forms an upper semilattice.
  % Kleene-Post 54 Degrees upper semi-lattice, jump an operator on degrees
  \cite{kleene54_upper_semi_lattic_degrees_recur_unsol}
\end{thm}
\begin{lem}
 $ \mathcal{D}$ does not form a lattice.
  \cite{kleene54_upper_semi_lattic_degrees_recur_unsol}
\end{lem}

@@

% \subsection{Turing Degrees of Problems related to CFGs}
% \cite{REEDY197577}
